---
title: "Single-Cell RNA-seq Preprocessing and Quality Control"
author: "Manisha Kalsan"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
subtitle: A Worked Example Using the 10x Genomics PBMC 3k Dataset
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  fig.width = 8,
  fig.height = 5
)
```

# Introduction

Single-cell RNA sequencing (scRNA-seq) measures gene expression in individual
cells rather than averaging expression across a bulk tissue sample. Where
bulk RNA-seq returns one expression profile per sample, scRNA-seq returns one
profile *per cell*, typically for thousands to tens of thousands of cells in
a single experiment. This resolution makes it possible to identify cell
types and states that would be invisible in bulk data, characterize rare
populations, trace differentiation trajectories, and study cell-to-cell
heterogeneity within an otherwise "homogeneous" tissue.

That resolution comes at a cost. Each cell contributes only a small amount of
RNA, so the resulting count data are noisy, sparse, and subject to technical
artifacts that bulk RNA-seq does not have to contend with in the same way
(dropout, variable capture efficiency, ambient RNA contamination, doublets).
A large fraction of the analytical effort in any scRNA-seq workflow —
arguably the most consequential fraction, since errors here propagate into
every downstream result — goes into **preprocessing and quality control**:
deciding which barcodes represent real cells, which genes are usably
detected, and how to normalize the resulting count matrix so that
expression values are comparable across cells.

This document walks through that preprocessing stage end to end, using the
canonical **PBMC 3k** dataset from 10x Genomics as a worked example. Each
section pairs a short conceptual explanation of *why* a step is done with
the runnable R code that does it, so the document can be read either as a
tutorial on the concepts or as an executable pipeline (or both, which is the
intent of R Markdown).

## What this document covers

1. The structure of 10x Genomics data and how it is loaded into R
2. Construction and anatomy of a `Seurat` object
3. Data exploration: sparsity, distributions of library size and gene detection
4. Quality-control metrics, in particular mitochondrial and ribosomal content
5. Statistical vs. literature-based thresholding strategies for filtering
6. Cell- and gene-level filtering
7. Two normalization strategies — log-normalization and `SCTransform` — and
   how their outputs differ
8. Saving processed objects for downstream analysis (clustering, cell-type
   annotation, differential expression — not covered here)

This document intentionally stops at the end of preprocessing. Clustering,
dimensionality reduction, and cell-type annotation are separate stages built
on top of a properly QC'd and normalized object, and are natural candidates
for a follow-up document.

## About the dataset

**PBMC 3k** is a peripheral blood mononuclear cell (PBMC) dataset released
by 10x Genomics as a public demonstration dataset for their Chromium
single-cell 3' gene expression assay. It contains gene expression
measurements for approximately 2,700 single cells, sequenced on an Illumina
platform.

Because PBMCs are a mixture of well-characterized immune cell types (T
cells, B cells, NK cells, monocytes, dendritic cells) with known marker
genes, and because the dataset is small enough to process on a laptop, it
has become the de facto "hello world" dataset for benchmarking and teaching
single-cell workflows — most single-cell tutorials (including Seurat's own)
use it for exactly this reason.

Key properties of the raw data:

| Property | Description |
|---|---|
| Assay | 10x Genomics Chromium, 3' gene expression |
| Organism | Human |
| Tissue | Peripheral blood mononuclear cells (PBMCs) |
| Cells (raw) | ~2,700 |
| Format | 10x "filtered" matrix: `barcodes.tsv`, `genes.tsv`/`features.tsv`, `matrix.mtx` |
| Reference genome | hg19 |
| Source | [10x Genomics datasets page](https://www.10xgenomics.com/resources/datasets) |

## The 10x Genomics output format

A 10x Chromium run produces, after alignment and cell calling with
`cellranger`, a sparse count matrix distributed as three files:

- **`matrix.mtx`** — the count matrix itself, stored in MatrixMarket sparse
  format (only nonzero entries are written to disk, since the vast majority
  of gene × cell entries are zero)
- **`barcodes.tsv`** — the cell barcode corresponding to each column of the
  matrix (one barcode = one droplet =, ideally, one cell)
- **`genes.tsv` / `features.tsv`** — the gene identifiers corresponding to
  each row of the matrix

Storing the matrix in sparse form is not a minor implementation detail — it
is what makes single-cell data tractable at all. A dense matrix for 2,700
cells × ~32,000 genes would need to hold ~86 million entries; the sparse
representation only needs to store the nonzero ones, which, as we'll see
below, is a small fraction of that.

# Environment Setup

```{r libraries}
# Install required packages (run once; commented out for a reproducible knit)
# install.packages(c("Seurat", "dplyr", "ggplot2", "patchwork", "Matrix", "RColorBrewer", "reshape2"))
# install.packages("BiocManager")
# BiocManager::install(c("SingleCellExperiment", "scater"))

library(Seurat)              # Primary single-cell analysis toolkit
library(dplyr)                # Data manipulation
library(ggplot2)              # Plotting
library(patchwork)            # Combining ggplot objects into panels
library(Matrix)                # Sparse matrix handling
library(RColorBrewer)         # Color palettes
library(reshape2)             # Reshaping data frames for plotting

# Set a random seed for reproducibility. Several downstream steps (e.g.
# SCTransform's internal subsampling, later clustering/UMAP steps) are
# stochastic, so fixing the seed makes the analysis re-runnable.
set.seed(452681)
```

We use `Seurat`, the most widely adopted R package for single-cell analysis,
as the primary toolkit. `SingleCellExperiment`/`scater` (Bioconductor's
alternative container and toolkit) are mentioned in the original script as
optional companions but are not required for the Seurat-based workflow
below.

```{r directories}
# Project directory structure. Adjust base_dir to your own path before knitting.
base_dir <- "~/Downloads/scrnaseq"

if (!dir.exists(base_dir)) {
  dir.create(file.path(base_dir, "data"), recursive = TRUE)
  dir.create(file.path(base_dir, "results"), recursive = TRUE)
  dir.create(file.path(base_dir, "plots"), recursive = TRUE)
}

knitr::opts_knit$set(root.dir = base_dir)
```

# Data Acquisition and Loading

## Downloading the data

The PBMC 3k filtered gene-barcode matrix is publicly hosted by 10x Genomics.
It is downloaded once and cached locally rather than re-fetched on every
knit.

```{r download, eval=FALSE}
# Source: 10x Genomics public datasets
url <- "https://cf.10xgenomics.com/samples/cell/pbmc3k/pbmc3k_filtered_gene_bc_matrices.tar.gz"
download_path <- file.path(base_dir, "pbmc3k_filtered_gene_bc_matrices.tar.gz")

if (!file.exists(download_path)) {
  download.file(url, destfile = download_path, mode = "wb")
}

# Extract into the data/ directory
untar(download_path, exdir = file.path(base_dir, "data"))
```

## Loading a 10x matrix into R

`Seurat::Read10X()` reads the three-file 10x format described above and
assembles it into a single sparse matrix (an object of class `dgCMatrix`),
with gene symbols as row names and cell barcodes as column names.

```{r load_data}
pbmc.data <- Read10X(data.dir = file.path(base_dir, "data", "filtered_gene_bc_matrices", "hg19"))

cat("Dataset dimensions:", dim(pbmc.data), "\n")
cat("Data type:", class(pbmc.data), "\n")
cat("First few gene names:", head(rownames(pbmc.data)), "\n")
cat("First few barcodes:", head(colnames(pbmc.data)), "\n")
```

## Why sparsity matters

In a typical droplet-based scRNA-seq experiment, the majority of gene ×
cell entries in the count matrix are zero. This is partly biological (most
genes are not expressed in most cell types at any given time) and partly
technical — a phenomenon usually called **dropout**: a transcript that is
genuinely present in the cell but is not captured, reverse-transcribed, or
sequenced deeply enough to be detected. Distinguishing "true zero" from
"dropout zero" is impossible at the level of a single entry, and this
ambiguity is one of the central statistical challenges in scRNA-seq analysis
(it motivates, among other things, the imputation and modeling choices in
methods like `SCTransform`, discussed later).

```{r sparsity_check}
non_zero_pct <- round(100 * sum(pbmc.data > 0) / (nrow(pbmc.data) * ncol(pbmc.data)), 2)
cat("Percentage of non-zero entries in the raw matrix:", non_zero_pct, "%\n")
```

Storing this matrix in sparse format (rather than as a dense matrix) is what
keeps memory usage manageable — the object only stores the coordinates and
values of nonzero entries.

# Creating the Seurat Object

A `Seurat` object is the central data structure used throughout the
workflow. It bundles together the raw count matrix, per-cell metadata
(library size, number of genes detected, and anything else computed later
such as QC metrics or cluster assignments), and, as the analysis proceeds,
normalized data, scaled data, dimensionality-reduction embeddings, and
nearest-neighbor graphs — all addressable through one object rather than as
separate, easily-desynchronized variables.

At creation time, two basic filters are already applied:

- **`min.cells = 3`**: a gene is kept only if it is detected (count > 0) in
  at least 3 cells. Genes detected in fewer cells than this are almost
  always uninformative noise and needlessly inflate the feature space.
- **`min.features = 200`**: a cell (barcode) is kept only if at least 200
  genes are detected in it. Barcodes with very few detected genes are
  usually empty droplets or fragments of lysed cells rather than intact
  single cells.

These are coarse, standard defaults intended to remove obviously unusable
barcodes/genes before any more principled QC is applied — they are not a
substitute for the QC step in the next section.

```{r create_seurat}
pbmc <- CreateSeuratObject(
  counts = pbmc.data,
  project = "pbmc3k",
  min.cells = 3,
  min.features = 200
)

cat("Number of cells retained:", ncol(pbmc), "\n")
cat("Number of genes retained:", nrow(pbmc), "\n")
```

## Automatically computed metadata

On creation, Seurat automatically computes two metrics for every cell:

- **`nCount_RNA`** — total UMI (Unique Molecular Identifier) count for the
  cell, i.e. total library size. UMIs are short random sequences attached
  to each transcript molecule before amplification, which let Seurat
  distinguish independent transcript molecules from PCR duplicates of the
  same molecule — this is why scRNA-seq counts are described as UMI counts
  rather than raw reads.
- **`nFeature_RNA`** — number of distinct genes detected (count > 0) in the
  cell.

```{r inspect_metadata}
head(pbmc@meta.data)

summary(pbmc$nCount_RNA)
summary(pbmc$nFeature_RNA)
```

These two numbers are the foundation of the QC step below: cells with
implausibly low values of either are usually not intact single cells, and
cells with implausibly high values are often doublets (two or more cells
captured in a single droplet and sequenced together as one barcode).

# Exploring Single-Cell Data Characteristics

Before doing any filtering or normalization, it is worth building intuition
for what makes single-cell count data different from bulk RNA-seq data,
since every preprocessing decision downstream is a response to one of these
properties.

Single-cell count data are characterized by:

1. **Sparsity** — most gene/cell entries are zero (shown above)
2. **High cell-to-cell variability** — even cells of the same type differ in
   library size and detected gene count, partly biology, partly technical
   variation in capture efficiency
3. **Technical noise** — amplification bias, batch effects, and
   sequencing depth all affect the observed counts independently of true
   expression
4. **Dropout** — false zeros, as discussed above

```{r gene_sparsity}
gene_sparsity <- apply(pbmc.data, 1, function(x) sum(x == 0) / length(x))

cat("Average gene sparsity (fraction of zero entries):",
    round(mean(gene_sparsity), 3), "\n")

highly_sparse_genes     <- names(gene_sparsity[gene_sparsity > 0.95])
moderately_sparse_genes <- names(gene_sparsity[gene_sparsity > 0.5 & gene_sparsity <= 0.95])
lowly_sparse_genes      <- names(gene_sparsity[gene_sparsity <= 0.5])

cat("Highly sparse genes   (>95% zeros):", length(highly_sparse_genes), "\n")
cat("Moderately sparse genes (50-95%):",   length(moderately_sparse_genes), "\n")
cat("Lowly sparse genes    (<50% zeros):", length(lowly_sparse_genes), "\n")
```

Most genes fall into the "highly sparse" bucket. This is expected: any given
cell type only actively transcribes a subset of the genome, and PBMCs
comprise several distinct cell types, so genes specific to one lineage will
read as zero in all the others.

## Visualizing basic distributions

```{r basic_viz}
p1 <- ggplot(pbmc@meta.data, aes(x = nCount_RNA)) +
  geom_histogram(bins = 50, fill = "lightblue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of UMI Counts per Cell",
       x = "Total UMI Counts", y = "Number of Cells") +
  theme_minimal()

p2 <- ggplot(pbmc@meta.data, aes(x = nFeature_RNA)) +
  geom_histogram(bins = 50, fill = "lightcoral", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Genes Detected per Cell",
       x = "Number of Genes", y = "Number of Cells") +
  theme_minimal()

p3 <- ggplot(pbmc@meta.data, aes(x = nCount_RNA, y = nFeature_RNA)) +
  geom_point(alpha = 0.6, color = "darkgreen") +
  labs(title = "UMI Counts vs Genes Detected",
       x = "Total UMI Counts", y = "Number of Genes Detected") +
  theme_minimal()

p1
p2
p3

ggsave(file.path(base_dir, "plots", "umi_distribution.png"), p1, width = 8, height = 6)
ggsave(file.path(base_dir, "plots", "gene_distribution.png"), p2, width = 8, height = 6)
ggsave(file.path(base_dir, "plots", "counts_vs_genes.png"), p3, width = 8, height = 6)

print(p1)
print(p2)
print(p3)
```

The positive, roughly linear relationship between UMI counts and genes
detected (Plot 3) is expected under normal conditions — cells sequenced more
deeply will, on average, have more genes detected simply because there is
more opportunity to detect each one. Deviations from this relationship (a
cell with very high UMI counts but a disproportionately *low* number of
genes detected, for instance) are one of the signatures used to flag
doublets or damaged cells.

```{r save_raw}
saveRDS(pbmc, file.path(base_dir, "data", "pbmc3k.rds"))
```

---

# Quality Control and Preprocessing

With the raw object built and its general character understood, we now
compute the metrics used to decide which cells (and, later, which genes)
should be retained for downstream analysis.

## Why mitochondrial content matters

Cells that are dying, ruptured, or otherwise stressed tend to lose
cytoplasmic mRNA through membrane damage while retaining mitochondrial mRNA
somewhat longer, because mitochondria are membrane-bound organelles. The
practical consequence is that a **high proportion of a cell's transcripts
mapping to mitochondrial genes** is a strong signal of a low-quality or
dying cell rather than a genuine biological state (with a few
tissue-specific exceptions, e.g. cardiomyocytes, which are not relevant to
PBMCs). Human mitochondrial genes are conventionally identified by the
`MT-` gene symbol prefix.

Ribosomal gene content (`RPS`/`RPL` prefixes) is tracked as a secondary,
optional QC signal — it is more informative as a marker of cell-type-specific
translational activity than of cell viability, but unusually low or
unusually high values can still flag technical issues.

```{r qc_reload}
pbmc <- readRDS(file.path(base_dir, "data", "pbmc3k.rds"))
cat("Loaded PBMC dataset with", ncol(pbmc), "cells and", nrow(pbmc), "genes\n")
```

```{r qc_metrics}
# Percentage of counts mapping to mitochondrial genes
pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")

# Percentage of counts mapping to ribosomal genes (secondary QC signal)
pbmc[["percent.ribo"]] <- PercentageFeatureSet(pbmc, pattern = "^RP[SL]")

head(pbmc@meta.data)
```

## Visualizing QC metrics

Violin plots show the distribution of each metric across all cells; scatter
plots reveal relationships between metrics that a single-metric view would
miss (e.g., whether high mitochondrial content is concentrated among
low-library-size cells).

```{r qc_viz, fig.width=10}
p1 <- VlnPlot(pbmc, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
              ncol = 3, pt.size = 0.1)
p1

plot1 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "percent.mt") +
  labs(title = "UMI Counts vs Mitochondrial Percentage")

plot2 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "nFeature_RNA") +
  labs(title = "UMI Counts vs Genes Detected")

combined_scatter <- plot1 + plot2
combined_scatter

ggsave(file.path(base_dir, "plots", "qc_violin.png"), p1, width = 12, height = 6)
ggsave(file.path(base_dir, "plots", "qc_scatter.png"), combined_scatter, width = 12, height = 6)
```

## Setting filtering thresholds

There are two broad philosophies for choosing QC thresholds, and the code
below demonstrates both so they can be compared.

**1. Statistical / data-driven thresholding.** Compute a robust measure of
central tendency and spread for each metric — the median and the **median
absolute deviation (MAD)** rather than the mean and standard deviation,
because MAD is far less sensitive to the extreme outliers that scRNA-seq QC
metrics routinely contain — and flag cells beyond some multiple of MAD from
the median (3 MAD is a common convention, roughly analogous to a
z-score cutoff but robust to outliers).

**2. Literature-based / conventional thresholding.** Use fixed, commonly
cited cutoffs from the literature and from widely used tutorials (e.g. the
original Seurat PBMC3k vignette): `nFeature_RNA` between 200 and 2500, and
`percent.mt` below 20%. These are less tailored to the specific dataset but
are simple, well-precedented, and easy to communicate.

Neither approach is objectively "correct" — the data-driven approach adapts
to the specific sample but can be thrown off by a dataset that genuinely
contains many low-quality cells (inflating the MAD and loosening the
threshold accordingly); the literature-based approach is stable but was
calibrated on other datasets and assays. In practice, both are computed and
compared, and the final choice is a judgment call cross-checked against the
QC plots above.

```{r qc_summary_print}
cat("nFeature_RNA summary:\n")
print(summary(pbmc$nFeature_RNA))

cat("\nnCount_RNA summary:\n")
print(summary(pbmc$nCount_RNA))

cat("\npercent.mt summary:\n")
print(summary(pbmc$percent.mt))
```

```{r mad_thresholds}
# --- Statistical thresholds (median +/- 3 MAD) ---
feature_median <- median(pbmc$nFeature_RNA)
feature_mad    <- mad(pbmc$nFeature_RNA)
feature_lower  <- feature_median - 3 * feature_mad
feature_upper  <- feature_median + 3 * feature_mad

mt_median <- median(pbmc$percent.mt)
mt_mad    <- mad(pbmc$percent.mt)
mt_upper  <- mt_median + 3 * mt_mad

cat("Proposed filtering thresholds (median +/- 3 MAD):\n")
cat("nFeature_RNA: ", feature_lower, "-", feature_upper, "\n")
cat("percent.mt:   <", mt_upper, "\n")

# --- Literature-based / conventional thresholds ---
cat("\nAlternative thresholds (literature-based, Seurat PBMC3k convention):\n")
cat("nFeature_RNA: 200-2500\n")
cat("nCount_RNA:   1000-30000\n")
cat("percent.mt:   <20%\n")
```

## Applying the filters

We proceed with the literature-based thresholds, since they are the more
widely used convention for this exact dataset/assay combination and keep
the results comparable to other PBMC3k analyses.

```{r apply_cell_filter}
original_cell_count <- ncol(pbmc)

pbmc_filtered <- subset(
  pbmc,
  subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 20
)

filtered_cell_count <- ncol(pbmc_filtered)

cat("=== CELL FILTERING RESULTS ===\n")
cat("Original cell count: ", original_cell_count, "\n")
cat("Filtered cell count:  ", filtered_cell_count, "\n")
cat("Cells removed:        ", original_cell_count - filtered_cell_count, "\n")
cat("Percentage retained:  ",
    round(100 * filtered_cell_count / original_cell_count, 2), "%\n")
```

## Gene-level filtering

The `min.cells = 3` filter at object creation used the *unfiltered* cell
population. After removing low-quality cells, it is good practice to
re-check gene detection rates, since a gene that only appeared "detected in
enough cells" because of low-quality cells should be re-evaluated on the
cleaned population.

```{r gene_filter}
min_cells <- 3

counts_mat <- pbmc_filtered[["RNA"]]@layers[["counts"]]
genes_before <- nrow(counts_mat)

# Recover gene names on the counts layer if they were dropped (Seurat v5 Assay5 quirk)
if (is.null(rownames(counts_mat))) {
  rn <- tryCatch(Features(pbmc_filtered, assay = "RNA"), error = function(e) NULL)
  if (is.null(rn)) stop("No gene names found: rownames(counts) and Features() are both NULL.")
  rownames(counts_mat) <- rn
}

gene_counts <- Matrix::rowSums(counts_mat > 0)
keep_genes  <- names(gene_counts[gene_counts >= min_cells])

pbmc_filtered <- subset(pbmc_filtered, features = keep_genes)
genes_after <- length(keep_genes)

cat("=== GENE FILTERING RESULTS ===\n")
cat("Original gene count: ", genes_before, "\n")
cat("Filtered gene count:  ", genes_after, "\n")
cat("Genes removed:        ", genes_before - genes_after, "\n")
```

# Normalization

## Why normalize at all

Raw UMI counts are not directly comparable between cells, because two cells
expressing a gene at the same *true* level can produce very different raw
counts simply due to differences in sequencing depth (total UMIs captured
per cell). Normalization aims to remove this technical, depth-driven
variation so that differences in expression values between cells reflect
biology rather than how deeply each cell happened to be sequenced.

Two conceptually different normalization strategies are demonstrated here.

### Method 1 — Log-normalization

The traditional approach: scale each cell's counts by its total UMI count,
multiply by a constant scale factor (10,000 by convention, chosen simply to
bring values into a convenient range), and log-transform (`log1p`, i.e.
`log(1 + x)`, to handle zeros):

$$
x_{ij}^{\text{norm}} = \log\left(1 + \frac{x_{ij}}{\sum_j x_{ij}} \times 10{,}000\right)
$$

where \(x_{ij}\) is the raw count of gene \(i\) in cell \(j\). This is fast,
simple, and the historical default, but it treats every cell's total count
as an equally reliable estimate of sequencing depth and does not explicitly
model the count distribution's variance structure.

```{r log_normalize}
pbmc_log <- pbmc_filtered
pbmc_log <- NormalizeData(pbmc_log, normalization.method = "LogNormalize",
                          scale.factor = 10000)
cat("Log-normalization completed.\n")
```

### Method 2 — SCTransform

`SCTransform` takes a model-based approach: it fits a **regularized
negative binomial regression** of each gene's counts against sequencing
depth across all cells, pooling information across genes to regularize the
per-gene model parameters (since fitting each gene's model independently
would be noisy for lowly-expressed genes). The Pearson residuals from this
model — the deviation of each observed count from what the depth-based
model predicts — are used as the normalized expression values.

This has two practical advantages over log-normalization: it does not
require a separate downstream step to identify highly variable genes (the
model naturally highlights genes whose variance exceeds what sequencing
depth alone would predict), and it tends to better stabilize variance across
the full range of expression levels, including for lowly-expressed genes
where log-normalization is known to under- or over-correct. The tradeoff is
computational cost and, for some users, reduced interpretability of the
resulting values relative to a straightforward log ratio.

```{r sctransform}
pbmc_sct <- pbmc_filtered
pbmc_sct <- SCTransform(pbmc_sct, verbose = FALSE)
cat("SCTransform normalization completed.\n")
```

## Comparing the two methods

As a concrete illustration, we compare raw, log-normalized, and
SCTransform-normalized values for **CD3D**, a canonical marker of T cells
and one of the more variably expressed genes in a PBMC population — a good
stress test for how each method handles a gene with real biological
heterogeneity across cell types.

```{r compare_normalization}
if ("CD3D" %in% rownames(pbmc_filtered[["RNA"]])) {

  raw_counts <- GetAssayData(pbmc_filtered, assay = "RNA", layer = "counts")["CD3D", ]
  log_norm   <- GetAssayData(pbmc_log, assay = "RNA", layer = "data")["CD3D", ]
  sct_norm   <- pbmc_sct@assays$SCT@data["CD3D", ]

  norm_comparison <- data.frame(
    Raw         = as.numeric(raw_counts),
    LogNorm     = as.numeric(log_norm),
    SCTransform = as.numeric(sct_norm)
  )

  # Drop cells where the gene is undetected under every method
  norm_comparison <- norm_comparison[rowSums(norm_comparison) > 0, ]

  norm_long <- melt(norm_comparison)

  p_norm_comparison <- ggplot(norm_long, aes(x = variable, y = value, fill = variable)) +
    geom_boxplot() +
    labs(title = "CD3D Expression: Normalization Method Comparison",
         x = "Normalization Method", y = "Expression Value") +
    theme_minimal() +
    theme(legend.position = "none")

  p_norm_comparison

  ggsave(file.path(base_dir, "plots", "normalization_comparison.png"),
         p_norm_comparison, width = 8, height = 6)
}
```

Raw counts and the two normalized versions live on different scales by
construction, so the comparison above is about *shape* (spread, skew,
separation between expressing and non-expressing cells) rather than absolute
magnitude — a well-behaved normalization should compress the extreme
right-skew of the raw counts while still preserving the separation between
CD3D-high (T cell) and CD3D-low/absent (everything else) cells.

# Before/After Filtering Comparison

To make the effect of the QC filtering step visible rather than just
numeric, we recombine the pre- and post-filtering metadata into a single
data frame and re-plot the UMI-vs-gene relationship, colored by
mitochondrial percentage, faceted by filtering status.

```{r before_after_viz, fig.width=10}
pbmc$filter_status          <- "Original"
pbmc_filtered$filter_status <- "Filtered"

combined_metadata <- rbind(
  cbind(pbmc@meta.data[, c("nFeature_RNA", "nCount_RNA", "percent.mt")],
        Status = "Before Filtering"),
  cbind(pbmc_filtered@meta.data[, c("nFeature_RNA", "nCount_RNA", "percent.mt")],
        Status = "After Filtering")
)

p_before_after <- ggplot(combined_metadata,
                          aes(x = nCount_RNA, y = nFeature_RNA, color = percent.mt)) +
  geom_point(alpha = 0.6) +
  scale_color_gradient(low = "blue", high = "red", name = "% MT") +
  facet_wrap(~Status) +
  labs(title = "Quality Control: Before vs After Filtering",
       x = "Total UMI Counts", y = "Number of Genes Detected") +
  theme_minimal()

p_before_after

ggsave(file.path(base_dir, "plots", "before_after_filtering.png"),
       p_before_after, width = 12, height = 6)
```

The cells removed by filtering should visibly concentrate at the low end of
`nFeature_RNA`/`nCount_RNA` (likely empty droplets or debris) and at the
high-`percent.mt` end (likely dying or damaged cells) — if instead the
filtered-out cells look scattered randomly through the plot, that's a sign
the chosen thresholds may not be well matched to this dataset's actual
quality distribution and are worth revisiting.

# Saving Processed Outputs

Each object represents a different point in the preprocessing pipeline and
is saved separately so that downstream analyses (clustering, marker
detection, etc.) can start from whichever normalization is appropriate
without re-running QC from scratch.

```{r save_outputs}
saveRDS(pbmc_filtered, file.path(base_dir, "data", "pbmc_filtered.rds"))
saveRDS(pbmc_log,      file.path(base_dir, "data", "pbmc_log_normalized.rds"))
saveRDS(pbmc_sct,      file.path(base_dir, "data", "pbmc_sct_normalized.rds"))

qc_summary <- data.frame(
  Metric = c("Original Cells", "Filtered Cells", "Cells Removed", "Retention Rate (%)",
             "Original Genes", "Filtered Genes", "Genes Removed"),
  Value = c(original_cell_count, filtered_cell_count,
            original_cell_count - filtered_cell_count,
            round(100 * filtered_cell_count / original_cell_count, 2),
            genes_before, genes_after, genes_before - genes_after)
)

write.csv(qc_summary, file.path(base_dir, "results", "qc_summary.csv"), row.names = FALSE)

knitr::kable(qc_summary, caption = "Summary of QC filtering results")
```

```{r final_message, echo=FALSE}
cat("Key files created:\n")
cat("-", file.path(base_dir, "data", "pbmc_filtered.rds"), "\n")
cat("-", file.path(base_dir, "data", "pbmc_log_normalized.rds"), "\n")
cat("-", file.path(base_dir, "data", "pbmc_sct_normalized.rds"), "\n")
cat("-", file.path(base_dir, "plots"), "/*.png\n")
cat("-", file.path(base_dir, "results", "qc_summary.csv"), "\n")
```

# Summary and Next Steps

At this point the pipeline has produced three saved `Seurat` objects, all
built on the same QC-filtered set of cells and genes but carrying different
normalizations:

- `pbmc_filtered.rds` — QC-filtered, **not yet normalized** (raw counts only)
- `pbmc_log_normalized.rds` — log-normalized (`RNA` assay, `data` layer)
- `pbmc_sct_normalized.rds` — SCTransform-normalized (`SCT` assay)

Either normalized object is a valid starting point for the next stage of a
standard scRNA-seq workflow, which is not covered in this document:

1. **Feature selection** — identifying highly variable genes to focus
   downstream steps on the genes carrying the most biological signal
2. **Dimensionality reduction** — PCA (and, for visualization, UMAP/t-SNE)
   to compress the noisy, high-dimensional expression matrix into a
   tractable number of components
3. **Clustering** — grouping cells into putative cell populations based on
   similarity in the reduced-dimensional space (typically graph-based
   clustering, e.g. Louvain/Leiden on a shared-nearest-neighbor graph)
4. **Cell type annotation** — assigning biological identity to each cluster
   using canonical marker genes or reference-based label transfer
5. **Differential expression** — comparing expression between clusters or
   conditions to identify marker genes or condition-specific responses

Each of these steps depends critically on the QC and normalization
decisions made above — an under-filtered or poorly normalized input
propagates errors into every subsequent step — which is why this stage,
despite producing no biological "results" of its own, is treated as
foundational rather than a formality.

# Session Info

```{r session_info}
sessionInfo()
```
